---
output: 
  github_document:
    toc: TRUE
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
knitr::opts_chunk$set(message = F, warning = F)
```

### Toward declarative table generation

API describe layout of table (rows and cols) and compute (default to count records)

```{r, eval = F}
library(tidypivot)

ggtable(tidytitanic::passengers) |>
  set_cols(sex) |>
  set_rows(survived)
```

# {tidypivot} allows you to create tables by describing them (like ggplot2 plotting's description/declaration)

note: see original discussion here: https://evamaerey.github.io/mytidytuesday/2022-02-14-tables/tables.html and thoughtful contributions from @shannonpileggi and @brshallow https://github.com/EvaMaeRey/mytidytuesday/issues/3


> And, you know, I'd get a dataset. And, *in my head I could very clearly kind of picture*, I want to put this on the x-axis. Let's put this on the y-axis, draw a line, put some points here, break it up by this variable.  And then, like, getting that vision out of my head, and into reality, it's just really, really hard. Just, like, felt harder than it should be. Like, there's a lot of custom programming involved, where I just felt, like, to me, I just wanted to say, like, you know, *this is what I'm thinking, this is how I'm picturing this plot. Like you're the computer 'Go and do it'.*  ... and I'd also been reading about the Grammar of Graphics by Leland Wilkinson, I got to meet him a couple of times and ... I was, like, this book has been, like, written for me.   - https://www.trifacta.com/podcast/tidy-data-with-hadley-wickham/


# declarative table creation with ggplot2

### setup

```{r, out.width="49%", fig.show='hold'}
library(ggplot2)
StatSumLabel <- ggproto(`_class` = "StatSumLabel",
                        `_inherit` = StatSum,
                        default_aes = aes(label = after_stat(n)))

theme_grey(16) |>
  set_theme()
```

### declarative table build

```{r, out.width="49%", fig.show='hold'}
library(ggplot2)
# 1. Declare table form
tidytitanic::tidy_titanic |>
  ggplot() + 
  aes(x = sex, # cols
      y = survived)   # rows

# 2. Now fill in the table
last_plot() + 
  geom_text(stat = StatSumLabel)  # geom_count could be used too
```

## Status quo table creation: Harder than it should be?

1. grouping on vars anticipated to be rows and cols
2. compute
3. pivoting 

```{r}
library(dplyr)
library(tidyr)

tidytitanic::tidy_titanic |> 
  group_by(survived, 
           sex) |> 
  count() |> 
  pivot_wider(
    names_from = sex,
    values_from = n
              )
```


## Under the hood, the mechanics follow the status quo...

- group by rows and columns
- value in data to consider (1 if not specified)
- wt, weight the value (1 if not specified)
- fun - do an operation (on value) within group 


# So internal helpers `data_filter`, `data_define_value`, `data_to_grouped` etc...

<details>

```{r helpers}
data_filter <- function(data, filter){
  
  dplyr::filter(data, {{filter}})
  
}


data_define_value <- function(data, value = NULL, wt = NULL){
  
    value_quo <- rlang::enquo(value)
    wt_quo    <- rlang::enquo(wt)

      if(rlang::quo_is_null(value_quo) ){

      ## adding a value as 1 if there is none
      data <- data |>
        dplyr::mutate(value = 1)
      
    }else{
      
      data <- data |> 
          dplyr::mutate(value = {{value}})
        
    }
    
    #### weighting ####
    
    if(!rlang::quo_is_null(wt_quo) ){
     
      data <- data |>
        dplyr::mutate(value = .data$value * {{wt}}) 
   }
    
    data
  
}


data_to_grouped <- function(data, cols, rows){
  
    ### grouping by tabulation vars col and row
    data |>
      dplyr::group_by(dplyr::across(c({{cols}}, {{rows}})),
                      .drop = TRUE)
  
  
}


data_grouped_to_summarized <- function(data, fun = NULL){
  
      if(is.null(fun))  {fun <- sum}

    ## adding a value as 1 if there is none
  
    ### summarizing ####
    
    data |>
        dplyr::summarise(summary = fun(.data$value))
  
  
}


data_summarized_to_proportioned <- function(data, prop = F, percent = F, within = NULL, round = 2){ 
    # proportion case or percent
  
    within_quo        <- rlang::enquo(within)
    # totals_within_quo <- rlang::enquo(totals_within)

    if(is.null(prop)) {prop <- FALSE}

    if(prop|percent){
      
      mult <- ifelse(percent, 100, 1)
      if(is.null(round)){round <- ifelse(percent, 1, 3)}

      # prop is across all data
        if(rlang::quo_is_null(within_quo) ){

            data <- data |>
              dplyr::ungroup() |>
              dplyr::mutate(prop = round(.data$summary*mult/sum(.data$summary), round))

        # prop is within categories specified by within variable
        }else{

              data <- data |>
                dplyr::ungroup() |>
                dplyr::group_by(dplyr::across(c({{within}})),
                                .drop = FALSE) |>
                dplyr::mutate(prop = round(.data$summary*mult/sum(.data$summary), round))

        }
    }
    
  if(prop|percent){data$display <- data$prop}else{data$display <- data$summary}
  
  data

}


data_proportioned_to_pivoted <- function(data, pivot = TRUE, cols = NULL){
  
    cols_quo  <- rlang::enquo(cols)

    tidy <- data |>
      dplyr::ungroup()


    # do not pivot if argument pivot false or if no columns specified
    if(pivot == F){

      tidy 

      # otherwise pivot by columns
    }else
    
    
    if(rlang::quo_is_null(cols_quo) & pivot){
      
      tidy <- tidy |> dplyr::select(-summary)
      if(!is.null(data$prop)|!is.null(data$percent)){
        tidy <- tidy |>  dplyr::select(-prop)
      }
        
      tidy |>
        dplyr::rename(value = display)
      
    } else
      
    
    if(!rlang::quo_is_null(cols_quo) & pivot){
      # keep only display column, and tabulation vars
      tidy <- tidy |> dplyr::select(-summary)
      if(!is.null(data$prop)|!is.null(data$percent)){
        tidy <- tidy |>  dplyr::select(-prop)
      }

      tidy |>
        tidyr::pivot_wider(names_from = {{cols}}, values_from = display)

    }

  }
```


.


```{r}
tidytitanic::flat_titanic |> 
  data_filter(freq > 35) |>
  data_define_value(value = freq) |> 
  data_to_grouped(rows = survived, cols = sex) |>
  data_grouped_to_summarized() |>
  data_summarized_to_proportioned(percent = T, within = survived) |>
  data_proportioned_to_pivoted(cols = sex)
```




</details>

# Encapsulating steps in `pivotr()`


<details>

```{r pivotr}
#' @export
pivotr <- function(data,
                   rows = NULL,
                   cols = NULL,
                   
                   value = NULL,
                   wt = NULL,
                       
                   fun = NULL,
                       
                   filter = TRUE,
                   prop = FALSE,
                   percent = FALSE,
                   round = NULL,
                       
                   within = NULL,

                   pivot = TRUE
){

  data |> 
  data_filter({{filter}}) |>
  data_define_value(value = {{value}}, wt = {{wt}}) |> 
  data_to_grouped(rows = {{rows}}, cols = {{cols}}) |>
  data_grouped_to_summarized(fun = fun) |>
  data_summarized_to_proportioned(prop = prop, percent = percent, within = {{within}}, round = round) |>
  data_proportioned_to_pivoted(pivot = pivot, cols = {{cols}})
  
}

```


</details>

### Examples

```{r}

tidytitanic::flat_titanic |> 
  pivotr(value = freq, 
         rows = survived, cols = sex, 
         percent = T, within = survived)

tidytitanic::flat_titanic |> 
  pivotr(value = freq, rows = survived, 
         cols = sex, filter = sex == "Female")

```


```{r}
library(tidytitanic)

tidy_titanic |> pivotr()

tidy_titanic |> pivotr(rows = sex, cols = survived)

tidy_titanic |> pivotr(rows = c(sex, age), cols = survived)

tidy_titanic |> pivotr(rows = sex, cols = survived, pivot = F)

flat_titanic |> pivotr(rows = sex, value = freq, prop = TRUE)

flat_titanic |> pivotr(rows = sex, cols = survived, value = freq, prop = TRUE)

flat_titanic |> pivotr(rows = sex, cols = survived, value = freq, prop = TRUE, within = sex)
```

# toward a piped workflow 


## tidypivot object

<details>


https://evamaerey.github.io/mytidytuesday/2024-07-02-s3-tables/s3-tables-tidypivot.html


```{r tidypivot-object}
#' @export
new_tidypivot <- function(data = data.frame(),
                          rows = NULL,
                          cols = NULL,
                          value = NULL,
                          wt = NULL,
                          fun = NULL,
                          filter = TRUE,
                          prop = FALSE,
                          percent = FALSE,
                          round = NULL,
                          within = NULL,
                          pivot = TRUE) {

  # table specification components !
  tp_obj <- list(
    # data = data,
    # rows = rows,
    # cols = cols,
    # value = value,
    # wt = wt,
    # fun = fun,
    # prop = prop,
    # percent = percent,
    # round = round,
    # within = within,
    # pivot = pivot
    # # more 'slots' to be added
  )

  # declare class 'tidypivot'
  class(tp_obj) <- "tidypivot"

  # Return the created object
  invisible(tp_obj)

}

#' @export
print.tidypivot <- function(tp_obj){
  
  print(do.call(pivotr, tp_obj))
  
  invisible(tp_obj)
  
}
```

</details>

## pipe-friendly user facing functions


<details>


```{r ggtable}
#' @export
ggtable <- function(data = NULL){
  
  # thedata <<- data # don't love this
  data <- data %||% data.frame()
  
  tp_obj <- new_tidypivot()
  
  tp_obj$data <- data
  
  last_tp_obj <<- tp_obj
  
  tp_obj

}


#' @export
last_table <- function(){
  
  last_tp_obj
  
}

```


```{r}
# knitr::opts_chunk$set(results = "asis")
```

```{r}
ggtable() 


tidytitanic::tidy_titanic |> head()

ggtable(tidytitanic::tidy_titanic) 


last_table()
```



```{r set-rows}
#' @export
set_rows <- function(tp_obj, rows = NULL){
  
  tp_obj$rows <- enquo(rows)
  
  last_tp_obj <<- tp_obj
  
  tp_obj

  
}


#' @export
set_cols <- function(tp_obj, cols = NULL){
  
  tp_obj$cols <- enquo(cols)

  last_tp_obj <<- tp_obj
  
  tp_obj
  

}

#' @export
set_filter <- function(tp_obj, filter = TRUE){
  
  if(!filter){tp_obj$filter <- enquo(filter)}
  
  last_tp_obj <<- tp_obj
  
  tp_obj
  

}
```


```{r}
ggtable(tidytitanic::tidy_titanic) |>
  set_rows(sex) |>
  set_cols(survived) |>
  set_filter(TRUE)

# last_table |>
#   set_filter(sex == "Female")
```


```{r set-fun}
#' @export
set_fun <- function(tp_obj, fun = sum){

  tp_obj$fun <- fun

  last_tp_obj <<- tp_obj

  tp_obj
  
}


#' @export
set_fun_example <- function(tp_obj){

  tp_obj$fun <- function(x) sample(x, 1)


  last_tp_obj <<- tp_obj

  tp_obj
  
}


#' @export
set_fun_sample <- function(tp_obj, n = 2, sep = "; "){

  tp_obj$fun <- function(x) paste(sample(x, n, replace = F), collapse = sep)


  last_tp_obj <<- tp_obj

  tp_obj
  
}


#' @export
set_fun_list <- function(tp_obj, sep = "; "){

  tp_obj$fun <- function(x) paste(x, collapse = sep)


  last_tp_obj <<- tp_obj

  tp_obj
  
}


```



```{r set-value}
#' @export
set_value <- function(tp_obj, value = NULL){
  
  tp_obj$value <- enquo(value)

  last_tp_obj <<- tp_obj
  
  tp_obj
  
}



#' @export
set_wt <- function(tp_obj, wt = NULL){
  
  tp_obj$wt <- enquo(wt)

  last_tp_obj <<- tp_obj
  
  tp_obj
  
}


#' @export
set_weight <- function(tp_obj, weight = NULL){
  
  tp_obj$weight <- enquo(weight)
  
  last_tp_obj <<- tp_obj
  
  tp_obj
  
}
```


```{r set-prop}

#' @export
calc_prop <- function(tp_obj, within = NULL){
  
  tp_obj$percent <- FALSE
  tp_obj$prop <- TRUE
  tp_obj$within <- enquo(within)
  
  last_tp_obj <<- tp_obj
  
  tp_obj
  
}

#' @export
calc_percent <- function(tp_obj, within = NULL){
  
  tp_obj$prop <- FALSE
  tp_obj$percent <- TRUE
  tp_obj$within <- enquo(within)
  
  last_tp_obj <<- tp_obj
  
  tp_obj
  
}


#' @export
set_within <- function(tp_obj, within = NULL){
  
  tp_obj$within <- enquo(within)
  
  last_tp_obj <<- tp_obj
  
  tp_obj
  
}
```


```{r no-pivot}
#' @export
no_pivot <- function(tp_obj){
  
  tp_obj$pivot <- FALSE

  last_tp_obj <<- tp_obj

  tp_obj  
  
}

collect <- function(tp_obj){
  
  do.call(pivotr, tp_obj)
  
}
```

</details>

### Examples


```{r}
tidytitanic::passengers |> head()
```


```{r}
ggtable(tidytitanic::passengers) |>
  set_cols(sex) |>
  set_rows(survived)

last_table() |>
  calc_percent()

last_table() |>
  calc_prop()

last_table() |>
  calc_prop(within = sex)

# a null table...
ggtable(tidytitanic::flat_titanic) |>
  set_value(NA) |>
  set_rows(sex)

ggtable(tidytitanic::flat_titanic) |>
  set_wt(freq) |>
  set_rows(sex) |>
  set_cols(survived) 

last_table() |>
  set_fun(mean)

last_table() |> 
  no_pivot()


last_table() |>
  collect()  # returns what would be printed as data frame


ggtable(tidytitanic::flat_titanic) |>
  set_value(freq) |>
  set_rows(sex) |>
  set_fun_example()

ggtable(tidytitanic::flat_titanic) |>
  set_value(freq) |>
  set_rows(sex) |>
  set_fun_sample(n = 3)

ggtable(tidytitanic::flat_titanic) |>
  set_value(freq) |>
  set_rows(sex) |>
  set_fun_list()
```


# Packaging

```{r, eval = F}
knitrExtra::chunk_names_get()

knitrExtra::chunk_to_dir("helpers")
knitrExtra::chunk_to_dir("pivotr")
knitrExtra::chunk_to_dir("tidypivot-object")
knitrExtra::chunk_to_dir("ggtable")
knitrExtra::chunk_to_dir("set-rows")
knitrExtra::chunk_to_dir("set-fun")
knitrExtra::chunk_to_dir("set-value")
knitrExtra::chunk_to_dir("set-prop")
knitrExtra::chunk_to_dir("no-pivot")
```

```{r, eval = F}
devtools::document()
devtools::check()
devtools::install(pkg = ".", upgrade = "never") 
```


# ggplot2 extension case study...

```{r}
library(tidyverse)

ext_exports <- read_csv("https://raw.githubusercontent.com/EvaMaeRey/mytidytuesday/refs/heads/main/2024-11-19-gg-prefixes/exported_funs_exts_ggplot2_tidyverse_org.csv") |> 
  mutate(prefix = str_extract(fun_exported, ".*?_")) |> 
  mutate(prefix_long = str_extract(fun_exported, ".+_")) |> 
  mutate(ind_classic_prefix = prefix %in% c("stat_", "geom_", "theme_", "scale_", "coord_", "facet_"))


ggtable(ext_exports)

ggtable(ext_exports |> filter(ind_classic_prefix))

last_table() |>
  set_rows(user) 

last_table() |>
  set_cols(prefix)

last_table() |>
  set_rows(c(user, repo))

# last_table() |>
#   set_rows(c(user, repo))


read_csv("https://raw.githubusercontent.com/EvaMaeRey/mytidytuesday/refs/heads/main/2024-12-10-ggplot2-layer-composition/ggplot2_exported_layer_fun_composition.csv") |> 
  rename(prefix = fun_prefix) ->
ggplot2_layers_definers

ggplot2_layers_definers |>
  ggtable()

last_table() |>
  set_rows(type)

last_table() |>
  set_rows(type) |>
  set_cols(default_or_fixed) |>
  set_rows(c(prefix, type))
```





# examples/derivative

```{r}
knitr::opts_chunk$set(results = "markup")
```

Here are some examples where you might have derivative functions

```{r pivotr_countsummean}
pivot_count <- function(...){

  # maybe a wt version...
  pivotr(fun = length, ...) 

  }

pivot_average <- function(...){
  
  mean_na_rm <- function(x){mean(x, na.rm = T)}
  
  pivotr(fun = mean_na_rm, ...) 
  
}

pivot_sum <- function(...){
  
  pivotr(fun = sum, ...)
  
}

pivot_empty <- function(...){
  
nar <- function(x) return(NA)

  pivotr(fun = nar, ...)

}
```




```{r, eval = T}
library(magrittr)
library(tidytitanic)

passengers <- readr::read_csv("https://raw.githubusercontent.com/clauswilke/dviz.supp/master/data-raw/titanic/Titanic.csv")

head(passengers)

tidy_titanic |> pivot_count(rows = sex)
tidy_titanic |> pivot_count(rows = sex, col = survived)
flat_titanic |> pivot_sum(rows = survived, value = freq)
flat_titanic |> pivot_sum(rows = sex,  cols = survived, value = freq)

flat_titanic |> pivot_average(rows = sex,  cols = survived, value = freq)
flat_titanic |> pivot_empty(rows = survived, cols = age)

passengers |> pivot_average(rows = c(Sex, PClass), cols = Survived, value = Age)


```


# filling cells with examples from data.

```{r}
pivot_example <- function(...){

  sample1 <- function(x) sample(x, 1)
  pivotr(fun = sample1, ...)

}


pivot_samplen <- function(..., n = 3, sep = "; "){

  samplen <- function(x) paste(sample(x, n, replace = F), collapse = sep)

  pivotr(fun = samplen, ...) 

}

pivot_list <- function(..., sep = "; "){

  paste_collapse <- function(x) paste (x, collapse = sep)
  pivotr(fun = paste_collapse, ...) 
  
}
```


```{r, eval = T}
flat_titanic |> pivot_example(rows = sex, value = freq)

flat_titanic |> pivot_samplen(rows = sex, value = freq)

flat_titanic |> pivot_list(rows = sex, cols = survived, value = freq)

set.seed(12345)
passengers |> pivot_example(rows = Survived, cols = Sex, value = Name)
passengers |> pivot_samplen(rows = Survived, cols = Sex, value = Name, n = 2, sep = "; ") 

passengers |> pivot_samplen(rows = Survived, cols = Sex, value = Age, n = 7) 

passengers |> dplyr::sample_n(20) |> pivot_list(rows = Sex, cols = Survived, value = Age)

```


# proportions helpers
  
```{r}  
library(tidytitanic)
# pivot_prop
flat_titanic |> pivotr(rows = sex, 
                       value = freq, 
                       prop = TRUE) # pivot_prop

flat_titanic |> 
  pivotr(rows = sex, cols = survived, 
         value = freq, prop = TRUE)

flat_titanic |> 
  pivotr(rows = sex, cols = survived, 
         value = freq, prop = TRUE, within = sex)

# pivot_percent
flat_titanic |> 
  pivotr(rows = sex, cols = survived, 
         value = freq, percent = TRUE, within = sex)
```




```{r}
knitr::knit_exit()

```

