---
output: 
  github_document:
    toc: TRUE
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
knitr::opts_chunk$set(message = F, warning = F)
```


# {tidypivot} allows you to create tables by describing them

note: see original discussion here: https://evamaerey.github.io/mytidytuesday/2022-02-14-tables/tables.html and thoughtful contributions from @shannonpileggi and @brshallow https://github.com/EvaMaeRey/mytidytuesday/issues/3



## example


```{r}
library(tidyverse)
library(tidypivot)
options(scipen = 10)

ggplot2::diamonds %>% 
  pivot_helper(rows = cut, 
     cols = color)

tidytitanic::tidy_titanic %>% 
  pivot_helper(rows = sex,
               cols = survived,
               prop = T,
               within = sex)
```

### Internals

- group by rows and columns
- value in data to consider (1 if not specified)
- wt, weight the value (1 if not specified)
- fun - do an operation (on value) within group 

```{r tp, eval = T}
#' A function that pretty much let's you do anything in the tidypivot space, is carefully crafted, and is adapted to make the other functions easy to use
#'
# #' @inheritParams pivot_count
#'
#' @return
#' @export
#'
#' @examples
#' tidy_titanic %>% pivotr(rows = sex, cols = survived, fun = length) # pivot_count
#' flat_titanic %>% pivotr(rows = sex, value = freq, fun = mean) # pivot_calc
#' flat_titanic %>% pivotr(rows = sex, value = freq, fun = sum) # pivot_count (weighted sum)
#' nar <- function(x) return(NA)
#' flat_titanic %>% pivotr(rows = sex, cols = survived, fun = nar); #pivot_null
#' sample1 <- function(x) sample(x, 1)
#' flat_titanic %>% pivotr(rows = sex, cols = survived, fun = sample1, value = freq); #pivot_sample1
#' samplen <- function(x, n) paste(sample(x, 5, replace = F), collapse = ", ")
#' # flat_titanic %>% pivotr(rows = sex, cols = survived, fun = samplen, value = freq); #pivot_samplen
#' paste_collapse <- function(x) paste (x, collapse = ", ")
#' flat_titanic %>% pivotr(rows = sex, fun = paste_collapse, value = freq) # pivot_list
#' flat_titanic %>% pivotr(rows = sex, value = freq, prop = TRUE) # pivot_prop
#' flat_titanic %>% pivotr(rows = sex, cols = survived, value = freq, prop = TRUE)
#' flat_titanic %>% pivotr(rows = sex, cols = survived, value = freq, prop = TRUE, within = sex)
pivotr <- function(data,
                       rows = NULL,
                       cols = NULL,
                       value = NULL,
                       wt = NULL,
                       
                       fun = NULL,
                       
                       prop = FALSE,
                       percent = FALSE,
                       round = NULL,
                       
                       within = NULL,
                       withinfun = NULL,
                       
                       pivot = NULL,
                       wrap = NULL,
                       totals_within = NULL
){

  
    cols_quo          <- rlang::enquo(cols)
    value_quo         <- rlang::enquo(value)
    wt_quo            <- rlang::enquo(wt)
    within_quo        <- rlang::enquo(within)
    totals_within_quo <- rlang::enquo(totals_within)

    if(is.null(prop)) {prop <- FALSE}
    if(is.null(pivot)){pivot <- TRUE}
    if(is.null(wrap)) {wrap <- FALSE}

    if(is.null(fun))  {fun <- sum}

    ## adding a value as 1 if there is none
    
    if(rlang::quo_is_null(value_quo) ){

      data <- data %>%
        dplyr::mutate(value = 1)
      
    }else{
      
      data <- data %>% 
          dplyr::mutate(value = {{value}})
        
    }
    
    #### weighting ####
    
    if(!rlang::quo_is_null(wt_quo) ){
     
      data <- data %>%
        dplyr::mutate(value = value * {{wt}}) 
   }
    
    
    ### grouping by tabulation vars col and row
    grouped <- data %>%
      dplyr::group_by(dplyr::across(c({{cols}}, {{rows}})),
                      .drop = FALSE)
  
    ### summarizing ####
    
    summarized <- grouped %>%
        dplyr::mutate(value = 1) %>%
        dplyr::summarise(value = fun(value))

    # proportion case or percent
    if(prop|percent){
      
      mult <- ifelse(percent, 100, 1)
      if(is.null(round)){round <- ifelse(percent, 1, 3)}

      # prop is across all data
        if(rlang::quo_is_null(within_quo) ){

            summarized <- summarized %>%
              dplyr::ungroup() %>%
              dplyr::mutate(value = round(value*mult/sum(value), round))

        # prop is within categories specified by within variable
        }else{

              summarized <- summarized %>%
                dplyr::ungroup() %>%
                dplyr::group_by(dplyr::across(c({{within}})),
                                .drop = FALSE) %>%
                dplyr::mutate(value = round(value*mult/sum(value), round))

        }
    }


    arranged <- summarized

    ungrouped <- arranged %>%
      dplyr::ungroup()

    tidy <- ungrouped


    # do not pivot if argument pivot false or if no columns specified
    if(pivot == F | rlang::quo_is_null(cols_quo)){

      tidy
      # tidy %>%
      #   dplyr::rename(count = .data$value)

      # otherwise pivot by columns
    }else{

      tidy %>%
        tidyr::pivot_wider(names_from = {{cols}})

    }

  }

```


# examples/derivitive

```{r}

tidy_titanic %>% pivotr(rows = sex, cols = survived, fun = length) # pivot_count
flat_titanic %>% pivotr(rows = sex, value = freq, fun = mean) # pivot_calc
flat_titanic %>% pivotr(rows = sex, value = freq, fun = sum) # pivot_count (weighted sum)
nar <- function(x) return(NA)
flat_titanic %>% pivotr(rows = sex, cols = survived, fun = nar); #pivot_null
sample1 <- function(x) sample(x, 1)
flat_titanic %>% pivotr(rows = sex, cols = survived, fun = sample1, value = freq); #pivot_sam
samplen <- function(x, n) paste(sample(x, 5, replace = F), collapse = ", ")
# flat_titanic %>% pivotr(rows = sex, cols = survived, fun = samplen, value = freq); #pivot_s
paste_collapse <- function(x) paste (x, collapse = ", ")
flat_titanic %>% pivotr(rows = sex, fun = paste_collapse, value = freq) # pivot_list
flat_titanic %>% pivotr(rows = sex, value = freq, prop = TRUE) # pivot_prop
flat_titanic %>% pivotr(rows = sex, cols = survived, value = freq, prop = TRUE)
flat_titanic %>% pivotr(rows = sex, cols = survived, value = freq, prop = TRUE, within = sex)

```


---

```{r Tidypivot}
# library(R6)

# library(tidytitanic)
# pivot_helper(data = passengers, rows = sex, cols = survived)

Tidypivot <- R6::R6Class("Tidypivot",
                     public = list(

                       # objects
                       data = NULL,
                       rows = NULL,
                       cols = NULL,
                       fun = NULL,
                       value = NULL,
                       wt = NULL,
                       within = NULL,
                       withinfun = NULL,
                       pivot = NULL,
                       wrap = NULL,
                       out = NULL
                       ,

                       # functions
                       update = function(data = NULL, rows = NULL, cols = NULL,
                                         fun = NULL, value = NULL, wt = NULL, pivot_logical = NULL){ # a method

                         # updating

                         if(!is.null(data)){self$data <- data}
                         if(!is.null(rows)){self$rows <- rows}
                         if(!is.null(cols)){self$cols <- cols}
                         if(!is.null(fun)){self$fun <- fun}
                         if(!is.null(wt)){self$wt <- wt}
                         if(!is.null(value)){self$value <- value}
                         if(!is.null(pivot_logical)){self$pivot <- pivot_logical}

                         # displaying
                         self$out <- 'tidypivot::pivot_helper(data, rows, cols, fun, value, wt, pivot_logical)'

                         if(!is.null(self$data)) {self$out <- self$out %>% stringr::str_replace("data",  self$data)  }
                         if(!is.null(self$rows)) {self$out <- self$out %>% stringr::str_replace("rows",  self$rows)  }
                         if(!is.null(self$cols)) {self$out <- self$out %>% stringr::str_replace("cols",  self$cols)  }
                         if(!is.null(self$fun))  {self$out <- self$out %>% stringr::str_replace("fun",   self$fun)   }
                         if(!is.null(self$value)){self$out <- self$out %>% stringr::str_replace("value", self$value) }
                         if(!is.null(self$wt))   {self$out <- self$out %>% stringr::str_replace("wt",    self$wt)    }
                         if(!is.null(self$pivot)){self$out <- self$out %>% stringr::str_replace("pivot_logical", self$pivot) }

                         if(is.null(self$data)) {self$out <- self$out %>% stringr::str_replace("data",  "NULL")}
                         if(is.null(self$rows)) {self$out <- self$out %>% stringr::str_replace("rows",  "NULL")}
                         if(is.null(self$cols)) {self$out <- self$out %>% stringr::str_replace("cols",  "NULL")}
                         if(is.null(self$fun))  {self$out <- self$out %>% stringr::str_replace("fun",   "NULL")}
                         if(is.null(self$value)){self$out <- self$out %>% stringr::str_replace("value", "NULL")}
                         if(is.null(self$wt))   {self$out <- self$out %>% stringr::str_replace("wt",    "NULL")}
                         if(is.null(self$pivot)){self$out <- self$out %>% stringr::str_replace("pivot_logical", "NULL")}

                         invisible(self)          #returns

                       },

                       print = function() {  # print method; default is to print everything

                         # str(self)
                         self$out


                       }
                     )
)


#' Title
#'
#' @param data
#' @param rows
#' @param cols
#' @param fun
#' @param value
#' @param wt
#' @param pivot_logical
#'
#' @return
#' @export
#'
#' @examples
#' tp_init(data = "tidytitanic::tidy_titanic")
tp_init <- function(data = NULL, quietly = FALSE){

  if(exists("my_tp")){rm(my_tp)}

  # my_tp <- Tidypivot$new()
  my_tp <<- Tidypivot$new()


  my_tp$update(data)

  if(!quietly){
  eval(parse(text =  my_tp$out))
  }

}


tp_init_pipe <- function(data = NULL, quietly = FALSE){

  # if(exists("my_tp")){rm(my_tp)}

  my_tp_pipe <- Tidypivot$new()
  # my_tp <<- Tidypivot$new()


  my_tp_pipe$update(data)

  if(!quietly){
    eval(parse(text =  my_tp_pipe$out))
  }

}
#
#
#' Title
#'
#' @param nothing
#' @param data
#' @param rows
#' @param cols
#' @param fun
#' @param value
#' @param wt
#' @param pivot_logical
#'
#' @return
#' @export
#'
#' @examples
#'tp_init(data = "tidytitanic::tidy_titanic")
#'   tp_add(rows = "sex")
#'   tp_add(cols = "class")
#'
#'tp_init(data = "tidytitanic::tidy_titanic") |>
#'   tp_add(rows = "sex") |>
#'   tp_add(cols = "class")
#'
tp_add <- function(nothing = NULL, data = NULL, rows = NULL, cols = NULL,
                                   fun = NULL, value = NULL, wt = NULL,
                   pivot_logical = NULL, quietly = FALSE){



  my_tp$update(data,
               rows,
               cols,
               fun,
               value,
               wt,
               pivot_logical)

  if(!quietly){
  eval(parse(text =  my_tp$out))
  }

}


tp_add_pipe <- function(nothing = NULL, data = NULL, rows = NULL, cols = NULL,
                   fun = NULL, value = NULL, wt = NULL,
                   pivot_logical = NULL, quietly = FALSE){



  my_tp_pipe$update(data,
               rows,
               cols,
               fun,
               value,
               wt,
               pivot_logical)

  if(!quietly){
    eval(parse(text =  my_tp_pipe$out))
  }

}

#
#
#' Title
#'
#' @param vars
#'
#' @return
#' @export
#'
#' @examples
#'   tp_init(data = "tidytitanic::tidy_titanic")
#'
#' set_data(data = tidytitanic::tidy_titanic) %>%
#'  set_rows(age)
set_rows <- function(first, vars){

  var_names = deparse(substitute(vars))

  var_names
  tp_add(rows = deparse(substitute(vars)))

}


#
#
#' Title
#'
#' @param vars
#'
#' @return
#' @export
#'
#' @examples
#' tp_init(data = "tidytitanic::tidy_titanic")
#'
#' set_data(data = tidytitanic::tidy_titanic) |>
#' set_rows(age) |>
#' set_cols(c(survived, sex))
set_cols <- function(first, vars){

  var_names = deparse(substitute(vars))

  var_names
  tp_add(cols = var_names)

}


#' Title
#'
#' @param vars
#'
#' @return
#' @export
#'
#' @examples
#' set_data(tidytitanic::tidy_titanic)
#'
set_data <- function(data){

  my_tp <<- Tidypivot$new()
  return(my_tp)

  my_tp$update(data = data,
               rows,
               cols,
               fun,
               value,
               wt,
               pivot_logical)


  eval(parse(text =  my_tp$out))


  # tp_add(data = deparse(substitute(data)))

}


```





# Piping



```{r}
tp_init(data = "tidytitanic::tidy_titanic")

set_data(tidytitanic::tidy_titanic) |> 
   set_rows(sex) |>
   set_cols(survived) |>
   set_cols(c(survived, age))
```





```{r}
knitr::knit_exit()

```


```{r}
library(tidypivot)

tp_init(data = "tidytitanic::tidy_titanic"); str(my_tp)
   tp_add(rows = "sex"); str(my_tp)
   tp_add(cols = "survived"); str(my_tp)
   tp_add(cols = "c(survived, class)"); str(my_tp)
```


```{r}
tp_init_pipe(data = "tidytitanic::tidy_titanic") |> 
   tp_add(cols = "class")   ; str(my_tp)
```   
   
# Alternate interface

---

# Step 0. Some observations

## ggplot2: user needs to describe layout of table

you can make a visual pivot table in ggplot2; analyst job is to describe the form. How will it look

specify 3 things - start with visual layout

- specify x
- specify y
- specify count type geom

```{r}
tidy_titanic %>% 
  ggplot() + 
  aes(x = Sex, y = Survived) + 
  geom_jitter() + 
  geom_count(color = "blue")
```

---

## With existing pivot tools, description isn't so visual

- specify vars
- specify aggregation
- specify visual arrangement (names from?) <- this comes last

```{r slow} 
tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = Survived, 
              values_from = count)
```

---

# Step 1a. Make Functions to allow description of final table, pivot_count and pivot_calc

## x argument is horizontal elements (columns) and y is vertical elements (rows)

```{r}
pivot_count_script <- readLines("./R/pivot_count.R")
```


```{r count, code = pivot_count_script}

```

```{r}
pivot_calc_script <- readLines("./R/pivot_calc.R")
```

```{r calc, code = pivot_calc_script}

```


# Step 1b. Using those functions

```{r examples}
# rows and cols
tidy_titanic %>% 
  pivot_count(rows = Survived, cols = Sex) 

# cols only
tidy_titanic %>% 
   pivot_count(cols = Sex)

# rows only
tidy_titanic %>% 
  pivot_count(rows = Survived) 

# two rows and col
tidy_titanic %>% 
  pivot_count(rows = c(Survived, Class), cols = Sex)

# two rows and col and contains zero counts
tidy_titanic %>% 
  pivot_count(rows = c(Survived, Class), cols = c(Sex, Age))

# two rows and col and contains zero counts
tidy_titanic %>% 
  pivot_count(rows = c(Survived, Class), cols = c(Sex, Age), pivot = F)

# count all
tidy_titanic %>% 
   pivot_count()

# for fun organize like it will appear visually in code
tidy_titanic %>% 
  pivot_count(                          cols = Sex, 
              rows = c(Survived, Class)        )
```


## After examining your table you might actually want to have the calculation in long form (for use in something like ggplot2).  This is what pivot = F is for!

```{r}
tidy_titanic %>% 
  pivot_count(cols = Sex, rows = Survived, pivot = F)
```

## 1b. pivot_calc using pivot calc function for non count aggregation

### just for fun arrange the code how the table will look

```{r calc2}
flat_titanic %>%
  pivot_calc(              cols = Sex, 
             rows = Survived, value = Freq, fun = sum)

flat_titanic %>% 
  pivot_count(cols = Sex, 
             rows = Survived, wt = Freq)

```

Issue: For this case, we should probably use pivot_count and allow for a wt argument.

## 1b style. use another tool to style

### goal of functions is not to style - just to make calculation faster by using a visually driven API

```{r style}
tidy_titanic %>%  
  pivot_count(cols = Sex, rows = c(Survived, Class)) %>% 
  group_by(Class) %>% 
  gt::gt()
```


```{r style2}
tidy_titanic %>% 
  pivot_count(cols = Sex, rows = c(Survived, Class, Age)) %>% 
  group_by(Age) %>% 
  gt::gt()
```

# Back to Step 0, Observations: use existing tools to calculate *proportions* is many step process

### feels like lots of gymnastics... a vis first approach is what we are after

```{r proportionslow}
tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(value = n()) %>% 
  group_by(Sex) %>% 
  mutate(prop = value/sum(value)) %>% 
  select(-value) %>% 
  pivot_wider(values_from = prop, names_from = Sex)
```

# Step 2a. build a function where visual arrangement leads.


```{r}
pivot_prop_script <- readLines("./R/pivot_prop.R")
```



```{r propfunction, code = pivot_prop_script}

```

# Step 2b. using the pivot_prop

```{r propusage}
tidy_titanic %>% 
  pivot_prop(rows = Survived, cols = Class, within = Class)

tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = Survived)

tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = c(Survived, Sex))
```

```{r propgraph}
tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = Survived, pivot = F) %>% 
  ggplot() +
  aes(x = Class, y = Sex) +
  facet_grid(rows = vars(Survived)) +
  geom_tile() +
  aes(fill = prop) + 
  geom_text(aes(label = prop %>% round(3)))
```

```{r propgraph2}
tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = c(Survived, Sex), pivot = F) %>% 
  ggplot() +
  aes(x = Class, y = 1) +
  facet_grid(rows = vars(Survived, Sex)) +
  geom_tile() +
  aes(fill = prop) + 
  geom_text(aes(label = prop %>% round(3)))

```

# Reflections, questions, issues

1. Does this already exist?
2. ~~How can API improve? possibly rows = vars(y00, y0, y), cols = vars(x).  and within = vars(?, ?)  This requires more digging into tidy eval.  What about multiple x vars?~~ These changes implemented thanks to Brian and Shannon
3. ~~How can internals improve?  Also tidy eval is old I think. defaults for missing data.~~ Using new {{}} tidy eval within and across, and rlang::quo_is_null() thanks to Brian
4. What about summary columns, rows.  Column totals, etc.  Maybe not very tidy... maybe allowed w/ error message.
5. Ideas about different API - more like ggplot2, where you would specify new dimensions of your table after piping.  Would require function to create non-data frame type object. Not sure about consistency dplyr/tidyr tools.  These just return dataframes/tibble.  I think being consistent with that expectation might be a good thing. Also less challenging to implement.


```{r, echo = F}
library(ggstamp)
set.seed(128790)
ggcanvas() + 
  stamp_tile(x = rnorm(500)/2.5, 
             y= rnorm(500)/2.5, 
             fill = "darkslateblue",
             alpha = runif(500, .25, 1),
             width = rnorm(500)/3,
             height = rnorm(500)/3, 
             color = alpha("black", .75)) + 
  stamp_point(size = 8, color = "grey85") + 
  stamp_spoke(angle = pi, size = 1.7, 
              radius = .65, color = "grey85") +
  stamp_point(size = 5, color = "grey85") + 
  stamp_point(size = 3, color = "grey55") + 

  stamp_polygon_holes(y0 = .25) + 
  stamp_text(label = "tidy", hjust = 1,
             color = "violetred", 
               vjust = -.2, size = 17) + 
  stamp_text(label = "pivot", hjust = -.1,
             angle = 0:25, color = "violetred", 
               vjust = -.2, size = 17, alpha = c(0:25/400)) + 
  stamp_spoke(angle = .45*0:25/25, size = 1.7, 
              radius = .85, color = "grey55", alpha = c(0:25/400)) +
  stamp_spoke(angle = .45, size = 1.7, 
              radius = .85, color = "grey55") +
  stamp_text(label = "pivot", hjust = -.1,
             angle = 25, color = "violetred", 
               vjust = -.2, size = 17) +
  ggstamp::theme_void_fill(fill = "darkslateblue") + 
  stamp_polygon(color = "lightslateblue", 
                alpha = 0, y0 = .25, size = 4)
```


# Other work in this space

- janitor::tably
- pivottabler




---

# And back again...

How to create a scatterplot with a pivot table

```{r}
library(ggplot2)
anscombe %>% 
  ggplot() + 
  aes(x = rank(x1), y = rank(y1)) + 
  geom_point() + 
  scale_y_continuous(breaks = 1:12 -.5) + 
  scale_x_continuous(breaks = 1:12 -.5) +
  theme(panel.grid.minor = element_blank())
# or

anscombe %>% 
  pivot_count(cols = x1, rows = y1) %>% 
  dplyr::arrange(-y1) %>% 
  knitr::kable()
```

```{r}
"tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = Survived, 
              values_from = count)" %>% 
  writeLines(con = tempfile()) 


ggstamp::ggdraft() + 
  ggstamp::stamp_text(label = "tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = Survived, 
              values_from = count)",
  hjust = 0, size = 8, x = .01, y = .4)

ggstamp::ggcanvas() + 
  ggstamp::stamp_circle(x0y0 = ggstamp::pos_polygon(n = 3),
                        fill = "green", alpha = .2, radius = 1.3) +
  ggstamp::stamp_text(xy = ggstamp::pos_polygon(n = 3, radius = 1.5),
                      label = c("ggplot2", "tidyr", "dplyr"))


```



```{r}
pivot_helper <- function(data,
                         
                       rows = NULL,
                       cols = NULL,
                       
                       value = NULL,
                       fun = NULL,
                       wt = NULL,
                       
                       prop = FALSE,
                       percent = FALSE,
                       round = NULL,
                       within = NULL,
                       withinfun = NULL,
                       
                       pivot = NULL,
                       wrap = NULL,
                       totals_within = NULL){

    cols_quo <- rlang::enquo(cols)
    value_quo <- rlang::enquo(value)
    wt_quo <- rlang::enquo(wt)
    within_quo <- rlang::enquo(within)
    totals_within_quo <- rlang::enquo(totals_within)

    if(is.null(prop)) {prop <- FALSE}
    if(is.null(pivot)){pivot <- TRUE}
    if(is.null(wrap)) {wrap <- FALSE}
    if(is.null(fun))  {fun <- sum   }

    ########## group data by all row and col vars #############
    grouped <- data %>%
      dplyr::group_by(dplyr::across(c({{cols}}, {{rows}})),
                      .drop = FALSE)

    ######### define value on which to summarize  ###############
    if(rlang::quo_is_null(value_quo) ){

      valued <- grouped %>%
        dplyr::mutate(value = 1)

    }else{

      valued <- grouped %>%
        dplyr::mutate(value = {{value}})

    }

    ######### weighting value if required ###################
    if(!rlang::quo_is_null(wt_quo)){

      valued <- valued %>%
        dplyr::mutate(value = value*{{wt}})

    }

    ######### summarizing by group according to function ########

    summarized <- valued %>%
      dplyr::summarise(summary = fun(value))


    ######## custom summaries props and percent, where within is defined ####

      ## proportion ####
    if(prop|percent){

      if(is.null(round)&prop){round = 3}
      if(is.null(round)&percent){round = 1}
      multiplier <- ifelse(percent, 100, 1)

      # proportion across all data, when 'within' is not defined
      if(rlang::quo_is_null(within_quo) ){

            summarized <- valued %>%
              dplyr::ungroup() %>%
              dplyr::summarise(summary = round(value*multiplier/sum(value), round))

            }else{

              summarized <- valued %>%
                dplyr::ungroup() %>%
                dplyr::group_by(dplyr::across(c({{within}})),
                                .drop = FALSE) %>%
                dplyr::summarise(summary = round(value*multiplier/sum(value), round))

            }
    }


    ###  Just saving a place for adding rearrangement variable

    arranged <- summarized

    #####  Ungrouping
    ungrouped <- arranged %>%
      dplyr::ungroup()

    #############  yields tidy data
    tidy <- ungrouped


    ####### should we actually pivot the data ####
    # do not pivot if argument pivot false or if no columns specified
    if(pivot == F | rlang::quo_is_null(cols_quo)){

      out <- tidy  # for unpivoted value should be renamed conditionally

      }else{

      out <- tidy %>%
        tidyr::pivot_wider(names_from = {{cols}})

      }

    return(out)

}

```

