---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r}
knitr::opts_chunk$set(message = F, warning = F)
```


# Step 00. prep some data, records and flat

```{r prep}
library(tidyverse)
library(magrittr)

Titanic %>% 
  data.frame() %>% 
  uncount(weights = Freq) ->
tidy_titanic ; tidy_titanic %>% head()

Titanic %>% 
  data.frame() ->
flat_titanic ; flat_titanic %>% head()
```

---

# Step 0. Some observations

## ggplot2: user needs to describe layout of table

you can make a visual pivot table in ggplot2; analyst job is to describe the form. How will it look

specify 3 things - start with visual layout

- specify x
- specify y
- specify count type geom

```{r}
tidy_titanic %>% 
  ggplot() + 
  aes(x = Sex, y = Survived) + 
  geom_jitter() + 
  geom_count(color = "blue")
```

---

## With existing pivot tools, description isn't so visual

- specify vars
- specify aggregation
- specify visual arrangement (names from?) <- this comes last

```{r slow} 
tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(value = n()) %>% 
  pivot_wider(names_from = Survived)
```

---

# Step 1a. Make Functions to allow description of final table, pivot_count and pivot_calc

## x argument is horizontal elements (columns) and y is vertical elements (rows)

```{r}
pivot_count_script <- readLines("./R/pivot_count.R")
```


```{r count, code = pivot_count_script}

```

```{r}
pivot_calc_script <- readLines("./R/pivot_calc.R")
```

```{r calc, code = pivot_calc_script}

```


# Step 1b. Using those functions

```{r examples}
tidy_titanic %>% 
   pivot_count(x = Sex)

tidy_titanic %>% 
  pivot_count(y = Survived, x = Sex) 

tidy_titanic %>% 
  pivot_count(y = Survived, x = Sex, y0 = Class)

# for fun organize like it will appear visually in code
tidy_titanic %>% 
  pivot_count(                          x = Sex, 
              y0 = Class, y = Survived          )
```


## After designing your table you might actually want to have the calculation in long form.

```{r}
tidy_titanic %>% 
  pivot_count(x = Sex, y = Survived, pivot = F)
```

## 1b. pivot_calc using pivot calc function for non count aggregation

### just for fun arrange the code how the table will look

```{r calc2}
flat_titanic %>%
  pivot_calc(              x = Sex, 
             y = Survived, value = Freq, fun = sum)
```

## 1b style. use another tool to style

### goal of functions is not to style - just to make calculation faster by using a visually driven API

```{r style}
tidy_titanic %>% 
  pivot_count(x = Sex, y = Survived, y0 = Class) %>% 
  group_by(Class) %>% 
  gt::gt()
```

# 

```{r style2}
tidy_titanic %>% 
  pivot_count(x = Sex, y = Survived, y0 = Class, y00 = Age) %>% 
  group_by(Age) %>% 
  gt::gt()
```

# Back to Step 0, Observations: use existing tools to calculate *proportions* is many step process

### feels like lots of gymnastics... a vis first approach is what we are after

```{r proportionslow}
tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(value = n()) %>% 
  group_by(Sex) %>% 
  mutate(prop = value/sum(value)) %>% 
  select(-value) %>% 
  pivot_wider(values_from = prop, names_from = Sex)
```

# Step 2a. build a function where visual arrangement leads.


```{r}
pivot_prop_script <- readLines("./R/pivot_prop.R")
```



```{r propfunction, code = pivot_prop_script}

```

# Step 2b. using the pivot_prop

```{r propusage}
tidy_titanic %>% 
  pivot_prop(y = Survived, x = Class, within = Class)

tidy_titanic %>% 
  pivot_prop(y0 = Survived, 
               y = Sex, 
               x = Class, 
               within = Survived)

tidy_titanic %>% 
  pivot_prop(y0 = Survived, 
               y = Sex, 
               x = Class, 
               within = Survived,
               within2 = Sex)
```

```{r propgraph}
tidy_titanic %>% 
  pivot_prop(y0 = Survived, 
               y = Sex, 
               x = Class, 
               within = Survived, pivot = F) %>% 
  ggplot() +
  aes(x = Class, y = Sex) +
  facet_grid(rows = vars(Survived)) +
  geom_tile() +
  aes(fill = prop) + 
  geom_text(aes(label = prop %>% round(3)))
```

```{r propgraph2}
tidy_titanic %>% 
  pivot_prop(y0 = Survived, 
               y = Sex, 
               x = Class, 
               within = Survived, 
               within2 = Sex, pivot = F) %>% 
  ggplot() +
  aes(x = Class, y = 1) +
  facet_grid(rows = vars(Survived, Sex)) +
  geom_tile() +
  aes(fill = prop) + 
  geom_text(aes(label = prop %>% round(3)))

```

# Reflections and questions

1. Does this already exist?
2. How can API improve? possibly rows = vars(y00, y0, y), cols = vars(x).  and within = vars(?, ?)  This requires more digging into tidy eval.  What about multiple x vars?
3. How can internals improve?  Also tidy eval is old I think. defaults for missing data. 
4. What about summary columns, rows.  Column totals, etc.  Maybe not very tidy...


```{r, echo = F}
library(ggstamp)

ggcanvas() + 
  stamp_point(size = 8, color = "grey85") + 
  stamp_spoke(angle = pi, size = 1.7, 
              radius = .65, color = "grey85") +
  stamp_point(size = 5, color = "grey85") + 
  stamp_point(size = 3, color = "grey55") + 
  stamp_spoke(angle = .45, size = 1.7, 
              radius = .85, color = "grey55") +
  stamp_polygon_holes(y0 = .25) + 
  stamp_text(label = "tidy", hjust = 1.3,
             color = "violetred", 
               vjust = -.2, size = 17) + 
  stamp_text(label = "pivot", hjust = -.1,
             angle = 25, color = "violetred", 
               vjust = -.2, size = 17) + 
  ggstamp::theme_void_fill(fill = "darkslateblue") + 
  stamp_polygon(color = "lightslateblue", 
                alpha = 0, y0 = .25, size = 4)
```
