---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
knitr::opts_chunk$set(message = F, warning = F)
```



```{r}
library(tidypivot)
```



note: see original discussion here: https://evamaerey.github.io/mytidytuesday/2022-02-14-tables/tables.html and thoughtful contributions from @shannonpileggi and @brshallow https://github.com/EvaMaeRey/mytidytuesday/issues/3

# Step 00. prep some data, records and flat data frame

```{r prep}
library(tidyverse)
library(magrittr)

Titanic %>% 
  data.frame() %>% 
  uncount(weights = Freq) ->
tidy_titanic ; tidy_titanic %>% head()

Titanic %>% 
  data.frame() ->
flat_titanic ; flat_titanic %>% head()
```

---

# Step 0. Some observations

## ggplot2: user needs to describe layout of table

you can make a visual pivot table in ggplot2; analyst job is to describe the form. How will it look

specify 3 things - start with visual layout

- specify x
- specify y
- specify count type geom

```{r}
tidy_titanic %>% 
  ggplot() + 
  aes(x = Sex, y = Survived) + 
  geom_jitter() + 
  geom_count(color = "blue")
```

---

## With existing pivot tools, description isn't so visual

- specify vars
- specify aggregation
- specify visual arrangement (names from?) <- this comes last

```{r slow} 
tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = Survived, 
              values_from = count)
```

---

# Step 1a. Make Functions to allow description of final table, pivot_count and pivot_calc

## x argument is horizontal elements (columns) and y is vertical elements (rows)

```{r}
pivot_count_script <- readLines("./R/pivot_count.R")
```


```{r count, code = pivot_count_script}

```

```{r}
pivot_calc_script <- readLines("./R/pivot_calc.R")
```

```{r calc, code = pivot_calc_script}

```


# Step 1b. Using those functions

```{r examples}
# rows and cols
tidy_titanic %>% 
  pivot_count(rows = Survived, cols = Sex) 

# cols only
tidy_titanic %>% 
   pivot_count(cols = Sex)

# rows only
tidy_titanic %>% 
  pivot_count(rows = Survived) 

# two rows and col
tidy_titanic %>% 
  pivot_count(rows = c(Survived, Class), cols = Sex)

# two rows and col and contains zero counts
tidy_titanic %>% 
  pivot_count(rows = c(Survived, Class), cols = c(Sex, Age))

# two rows and col and contains zero counts
tidy_titanic %>% 
  pivot_count(rows = c(Survived, Class), cols = c(Sex, Age), pivot = F)

# count all
tidy_titanic %>% 
   pivot_count()

# for fun organize like it will appear visually in code
tidy_titanic %>% 
  pivot_count(                          cols = Sex, 
              rows = c(Survived, Class)        )
```


## After examining your table you might actually want to have the calculation in long form (for use in something like ggplot2).  This is what pivot = F is for!

```{r}
tidy_titanic %>% 
  pivot_count(cols = Sex, rows = Survived, pivot = F)
```

## 1b. pivot_calc using pivot calc function for non count aggregation

### just for fun arrange the code how the table will look

```{r calc2}
flat_titanic %>%
  pivot_calc(              cols = Sex, 
             rows = Survived, value = Freq, fun = sum)

flat_titanic %>% 
  pivot_count(cols = Sex, 
             rows = Survived, wt = Freq)

```

Issue: For this case, we should probably use pivot_count and allow for a wt argument.

## 1b style. use another tool to style

### goal of functions is not to style - just to make calculation faster by using a visually driven API

```{r style}
tidy_titanic %>%  
  pivot_count(cols = Sex, rows = c(Survived, Class)) %>% 
  group_by(Class) %>% 
  gt::gt()
```


```{r style2}
tidy_titanic %>% 
  pivot_count(cols = Sex, rows = c(Survived, Class, Age)) %>% 
  group_by(Age) %>% 
  gt::gt()
```

# Back to Step 0, Observations: use existing tools to calculate *proportions* is many step process

### feels like lots of gymnastics... a vis first approach is what we are after

```{r proportionslow}
tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(value = n()) %>% 
  group_by(Sex) %>% 
  mutate(prop = value/sum(value)) %>% 
  select(-value) %>% 
  pivot_wider(values_from = prop, names_from = Sex)
```

# Step 2a. build a function where visual arrangement leads.


```{r}
pivot_prop_script <- readLines("./R/pivot_prop.R")
```



```{r propfunction, code = pivot_prop_script}

```

# Step 2b. using the pivot_prop

```{r propusage}
tidy_titanic %>% 
  pivot_prop(rows = Survived, cols = Class, within = Class)

tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = Survived)

tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = c(Survived, Sex))
```

```{r propgraph}
tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = Survived, pivot = F) %>% 
  ggplot() +
  aes(x = Class, y = Sex) +
  facet_grid(rows = vars(Survived)) +
  geom_tile() +
  aes(fill = prop) + 
  geom_text(aes(label = prop %>% round(3)))
```

```{r propgraph2}
tidy_titanic %>% 
  pivot_prop(rows = c(Survived, Sex), 
               cols = Class, 
               within = c(Survived, Sex), pivot = F) %>% 
  ggplot() +
  aes(x = Class, y = 1) +
  facet_grid(rows = vars(Survived, Sex)) +
  geom_tile() +
  aes(fill = prop) + 
  geom_text(aes(label = prop %>% round(3)))

```

# Reflections, questions, issues

1. Does this already exist?
2. ~~How can API improve? possibly rows = vars(y00, y0, y), cols = vars(x).  and within = vars(?, ?)  This requires more digging into tidy eval.  What about multiple x vars?~~ These changes implemented thanks to Brian and Shannon
3. ~~How can internals improve?  Also tidy eval is old I think. defaults for missing data.~~ Using new {{}} tidy eval within and across, and rlang::quo_is_null() thanks to Brian
4. What about summary columns, rows.  Column totals, etc.  Maybe not very tidy... maybe allowed w/ error message.
5. Ideas about different API - more like ggplot2, where you would specify new dimensions of your table after piping.  Would require function to create non-data frame type object. Not sure about consistency dplyr/tidyr tools.  These just return dataframes/tibble.  I think being consistent with that expectation might be a good thing. Also less challenging to implement.


```{r, echo = F}
library(ggstamp)
set.seed(128790)
ggcanvas() + 
  stamp_tile(x = rnorm(500)/2.5, 
             y= rnorm(500)/2.5, 
             fill = "darkslateblue",
             alpha = runif(500, .25, 1),
             width = rnorm(500)/3,
             height = rnorm(500)/3, 
             color = alpha("black", .75)) + 
  stamp_point(size = 8, color = "grey85") + 
  stamp_spoke(angle = pi, size = 1.7, 
              radius = .65, color = "grey85") +
  stamp_point(size = 5, color = "grey85") + 
  stamp_point(size = 3, color = "grey55") + 

  stamp_polygon_holes(y0 = .25) + 
  stamp_text(label = "tidy", hjust = 1,
             color = "violetred", 
               vjust = -.2, size = 17) + 
  stamp_text(label = "pivot", hjust = -.1,
             angle = 0:25, color = "violetred", 
               vjust = -.2, size = 17, alpha = c(0:25/400)) + 
  stamp_spoke(angle = .45*0:25/25, size = 1.7, 
              radius = .85, color = "grey55", alpha = c(0:25/400)) +
  stamp_spoke(angle = .45, size = 1.7, 
              radius = .85, color = "grey55") +
  stamp_text(label = "pivot", hjust = -.1,
             angle = 25, color = "violetred", 
               vjust = -.2, size = 17) +
  ggstamp::theme_void_fill(fill = "darkslateblue") + 
  stamp_polygon(color = "lightslateblue", 
                alpha = 0, y0 = .25, size = 4)
```


# Other work in this space

- janitor::tably
- pivottabler


---

# And back again...

How to create a scatterplot with a pivot table

```{r}
library(ggplot2)
anscombe %>% 
  ggplot() + 
  aes(x = rank(x1), y = rank(y1)) + 
  geom_point() + 
  scale_y_continuous(breaks = 1:12 -.5) + 
  scale_x_continuous(breaks = 1:12 -.5) +
  theme(panel.grid.minor = element_blank())
# or

anscombe %>% 
  pivot_count(cols = x1, rows = y1) %>% 
  dplyr::arrange(-y1) %>% 
  knitr::kable()
```

```{r}
"tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = Survived, 
              values_from = count)" %>% 
  writeLines(con = tempfile()) 


ggstamp::ggdraft() + 
  ggstamp::stamp_text(label = "tidy_titanic %>% 
  group_by(Sex, Survived) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = Survived, 
              values_from = count)",
  hjust = 0, size = 8, x = .01, y = .4)

ggstamp::ggcanvas() + 
  ggstamp::stamp_circle(x0y0 = ggstamp::pos_polygon(n = 3),
                        fill = "green", alpha = .2, radius = 1.3) +
  ggstamp::stamp_text(xy = ggstamp::pos_polygon(n = 3, radius = 1.5),
                      label = c("ggplot2", "tidyr", "dplyr"))


```


